Clustering and Parameter Estimation for the Popularity Adjusted Block Model

### 2020-07-13 summary.Rmd
updates on writeup

---
title: "Clustering and Parameter Estimation for the Popularity Adjusted Block Model"
# output:
#   pdf_document:
#     citation_package: natbib
#     number_sections: yes
output: html_document
# geometry: "left=1cm,right=1cm,top=1cm,bottom=1.5cm"
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,vlined,ruled]{algorithm2e} 
- \setcitestyle{numbers,square}
bibliography: misc.bib
abstract: |
  In this paper, we connect two probabolistic models for graphs, the Popularity 
  Adjusted Block Model (PABM) and the Generalized Random Dot Product Graph 
  (GRDPG) and use properties established in this connection to aid in 
  clustering and parameter estimation. In particular, we note that the PABM 
  can be represented as latent positions such that points within the same 
  cluster lie on a subspace, and the subspaces that represent different 
  clusters are orthogonal to one another. Using this property as well as the 
  asymptotic properties of Adjacency Spectral Embedding of the GRDPG, we are 
  able to establish theoretical asymptotic results of our clustering and 
  parameter estimation methods for the PABM.
---

@@

@@

$$\begin{split}
  \min_{\theta, \phi, \psi, \tau} & ||\xi^{(1, 1)}||^2 + ||\xi^{(2, 2)}||^2 + ||\xi^{(1, 3)} - \xi^{(1, 4)}||^2 + ||\xi^{(2, 3)} + \xi^{(2, 4)}||^2 \\
  \text{s.t. } & \Xi = Z Q(\theta, \phi, \psi, \tau) \\
  & \xi^{(i, j)} \text{ is the } j^{th} \text{ column of } \Xi \text{ with rows from the } i^{th} \text{ cluster}
\end{split}$$

@@

@@

```{r}
results.df <- foreach(n = n.vec,.combine = dplyr::bind_rows) %do% {
  plyr::ldply(seq(iter), function(i) {
    Az <- draw.pabm.beta.2(n, a1, b1, a2, b2, alpha)
    A <- Az$A
    z <- Az$z
    clustering <- cluster.pabm(A, K)
    accuracy <- mean(clustering == z)
    if (accuracy < .5) accuracy <- 1 - accuracy
    dplyr::tibble(n = n, accuracy = accuracy) %>% 
      return()
  }, .parallel = TRUE) %>% 
    return()
}

results.df %>% 
  dplyr::group_by(n) %>% 
  dplyr::summarise(mean.acc = mean(accuracy),
                   sd.acc = sd(accuracy)) %>% 
  ggplot() + 
  labs(y = 'accuracy') + 
  geom_line(aes(x = n, y = mean.acc)) + 
  geom_errorbar(aes(x = n, ymin = mean.acc - sd.acc, ymax = mean.acc + sd.acc))
```

@@

>>>>

---
title: "Clustering and Parameter Estimation for the Popularity Adjusted Block Model"
output:
  pdf_document:
    citation_package: natbib
    number_sections: yes
# output: html_document
# geometry: "left=1cm,right=1cm,top=1cm,bottom=1.5cm"
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,vlined,ruled]{algorithm2e} 
- \setcitestyle{numbers,square}
bibliography: misc.bib
abstract: |
  In this paper, we connect two probabolistic models for graphs, the Popularity 
  Adjusted Block Model (PABM) and the Generalized Random Dot Product Graph 
  (GRDPG) and use properties established in this connection to aid in 
  clustering and parameter estimation. In particular, we note that the PABM 
  can be represented as latent positions such that points within the same 
  cluster lie on a subspace, and the subspaces that represent different 
  clusters are orthogonal to one another. Using this property as well as the 
  asymptotic properties of Adjacency Spectral Embedding of the GRDPG, we are 
  able to establish theoretical asymptotic results of our clustering and 
  parameter estimation methods for the PABM.
---

@@

theme_set(theme_bw())

@@

\begin{equation}
\begin{split}
  \min_{\theta, \phi, \psi, \tau} & ||\xi^{(1, 1)}||^2 + ||\xi^{(2, 2)}||^2 + ||\xi^{(1, 3)} - \xi^{(1, 4)}||^2 + ||\xi^{(2, 3)} + \xi^{(2, 4)}||^2 \\
  \text{s.t. } & \Xi = Z Q(\theta, \phi, \psi, \tau) \\
  & \xi^{(i, j)} \text{ is the } j^{th} \text{ column of } \Xi \text{ with rows from the } i^{th} \text{ cluster}
\end{split}
\end{equation}

@@

We will not consider clustering using the edge probability matrix $P$ since 
this will always result exact recovery of the original clusters.

@@

```{r fig.cap = 'IQR clustering error using Algorithm 1 for sample sizes from 64 to 1024. Simulations were repeated 100 times for each sample size.', cache = TRUE}
results.df <- foreach(n = n.vec,.combine = dplyr::bind_rows) %do% {
  plyr::ldply(seq(iter), function(i) {
    Az <- draw.pabm.beta.2(n, a1, b1, a2, b2, alpha)
    A <- Az$A
    z <- Az$z
    clustering <- cluster.pabm(A, K)
    error <- mean(clustering == z)
    if (error > .5) error <- 1 - error
    dplyr::tibble(n = n, error = error) %>% 
      return()
  }, .parallel = TRUE) %>% 
    return()
}

results.df %>% 
  dplyr::group_by(n) %>% 
  dplyr::summarise(med.err = median(error),
                   first.q = quantile(error, .25),
                   third.q = quantile(error, .75)) %>% 
  ggplot() + 
  labs(y = 'error') + 
  geom_line(aes(x = n, y = med.err)) + 
  geom_errorbar(aes(x = n, ymin = first.q, ymax = third.q))
```

We can also examine how the distribution of $(\hat{v}_i)^\top \hat{v_j}$ varies 
with $n$:

```{r, fig.cap = 'Between-cluster inner products of the eigenvectors of $A$ for varying sample sizes.', cache = TRUE}
densities.df <- plyr::ldply(n.vec, function(n) {
  Az <- draw.pabm.beta.2(n, a1, b1, a2, b2, alpha)
  A <- Az$A
  z <- Az$z
  n1 <- sum(z == 1)
  n2 <- sum(z == 2)
  V <- eigen(A, symmetric = TRUE)$vectors[, c(seq(3), n)]
  B <- V %*% t(V)
  inner.prods <- as.numeric(B[(n1+1):n, (n1+1):n])
  return(dplyr::tibble(n = n, inner.prods = inner.prods))
}, .parallel = TRUE)

ggplot(densities.df) + 
  geom_density(aes(x = inner.prods, colour = factor(n))) + 
  labs(x = 'between cluster inner products',
       colour = 'n') + 
  scale_colour_brewer(palette = 'Set1')
```

@@

```{r}
rot.mat.2 <- function(angle, axis = 1) {
  # rotation matrix for K=2

  if (axis %in% c(1, 3)) {
    rotation <- matrix(c(cos(angle), -sin(angle), sin(angle), cos(angle)),
                       nrow = 2, ncol = 2, byrow = TRUE)
    if (axis == 1) {
      return(as.matrix(Matrix::bdiag(1, rotation, -1)))
    } else {
      return(as.matrix(Matrix::bdiag(rotation, 1, -1)))
    }
  } else if (axis == 2) {
    return(matrix(c(cos(angle), 0, sin(angle), 0,
                    0, 1, 0, 0,
                    -sin(angle), 0, cos(angle), 0,
                    0, 0, 0, -1),
                  nrow = 4, ncol = 4,
                  byrow = TRUE))
  } else {
    stop(simpleError('axis must be 1, 2, or 3'))
  }
}

hyp.rot.mat.2 <- function(angle, axis = 1) {
  # hyperbolic rotation matrix for $K=2

  if (axis == 1) {
    return(matrix(c(1, 0, 0, 0,
                    0, 1, 0, 0, 
                    0, 0, cosh(angle), sinh(angle),
                    0, 0, sinh(angle), cosh(angle)), 
                  byrow = TRUE,
                  nrow = 4, ncol = 4))
  } else if (axis == 2) {
    return(matrix(c(1, 0, 0, 0,
                    0, cosh(angle), 0, sinh(angle),
                    0, 0, 1, 0,
                    0, sinh(angle), 0, cosh(angle)),
                  byrow = TRUE,
                  nrow = 4, ncol = 4))
  } else if (axis == 3) {
    return(matrix(c(cosh(angle), 0, 0, sinh(angle),
                    0, 1, 0, 0, 
                    0, 0, 1, 0, 
                    sinh(angle), 0, 0, cosh(angle)),
                  byrow = TRUE,
                  nrow = 4, ncol = 4))
  } else {
    stop(simpleError('axis must be 1, 2, or 3'))
  }
}

construct.Q <- function(angles) {
  rot.angles <- angles[1:3]
  hyp.angles <- c(angles[1], angles[4], -angles[4])
  rot.matrices <- lapply(seq(3), function(i) rot.mat.2(rot.angles[i], i))
  hyp.matrices <- lapply(seq(3), function(i) hyp.rot.mat.2(hyp.angles[i], i))
  matrices <- c(rot.matrices, hyp.matrices)
  return(Reduce(`%*%`, matrices))
}

obj.fun <- function(angles, Z.hat, clusters) {
  R <- construct.Q(angles)

  XU.c <- Z.hat %*% R
  in.prods <- XU.c[clusters == 1, ] %*% t(XU.c[clusters == 2, , drop = FALSE])
  return(
    sum(XU.c[clusters == 1, 1] ** 2) + 
      sum(XU.c[clusters == 2, 2] ** 2) + 
      sum((XU.c[clusters == 1, 3] + XU.c[clusters == 1, 4]) ** 2) + 
      sum((XU.c[clusters == 2, 3] - XU.c[clusters == 2, 4]) ** 2)
  )
}

transform.embedding <- function(Z.hat, clusters) {
  # init.angles <- runif(4, -pi, pi)
  init.angles <- rep(0, 4)
  angles.1 <- optim(init.angles, function(x) obj.fun(x, Z.hat, clusters), 
                    method = 'SANN')$par
  angles.2 <- optim(angles.1, function(x) obj.fun(x, Z.hat, clusters),
                    method = 'Nelder-Mead',
                    control = list(abstol = 1e-10, maxit = 1e3))$par
  R <- construct.Q(angles.2)
  return(Z.hat %*% R)
}
```

```{r}
n <- 64
z <- sample(c(1, 2), n, prob = c(alpha, 1 - alpha), replace = TRUE)
z <- sort(z)

X <- cbind(c(lambda11, rep(0, n2)),
           c(lambda12, rep(0, n2)),
           c(rep(0, n1), lambda21),
           c(rep(0, n1), lambda22))
Y <- cbind(c(lambda11, rep(0, n2)),
           c(rep(0, n1), lambda21),
           c(lambda12, rep(0, n2)),
           c(rep(0, n1), lambda22))
P <- X %*% t(Y)
A <- draw.graph(P)
Z <- embedding(P)
Zhat <- embedding(A, 3, 1)
```

Setting $n = 64$ and generating one example set of $\{\lambda^{(kl)}\}$, we can 
see the latent positions $XU$ follow the expected properties 
(Fig. \ref{fig:xu}).

```{r xu, fig.cap = 'Latent positions $XU$ used to construct the PABM edge probability matrix $P$ under the GRDPG framework. Note that the the points follow the properties outlined in Section 3.2.'}
perm <- cbind(c(1, 0, 0, 0),
              c(0, 0, 1, 0),
              c(0, 1, 0, 0),
              c(0, 0, 0, 1))
U <- eigen(perm)$vectors[, c(3, 1, 2, 4)]
XU <- X %*% U

pairs(XU, asp = 1, col = z)
```

If we instead take the ASE of $P = X \Pi X^\top$, we fail to obtain an 
embedding that has the same properties (Fig. \ref{fig:ase_of_p}).

```{r ase_of_p, fig.cap = 'The adjacency spectral embedding of the edge probability matrix $P$. Note that even though $P = (XU) I_{p, q} (XU)^\\top$, the resulting embedding does not follow the same properties of $XU$, even up to rotation.'}
pairs(Z, asp = 1, col = z)
```

Denoting $Z$ as the ASE of $P$, we can recover $ZQ$ that is equivalent to $XU$ 
(Fig. \ref{fig:q_trans}).

```{r q_trans, fig.cap = 'Pair plots of $ZQ$ where $Q$ was found by minimizing objective function (1). This embedding is equivalent to $XU$ up to a 90-degree rotation.'}
XU.prime <- transform.embedding(Z, z)
pairs(XU.prime, col = z, asp = 1)
```

The ASE of the adjacency matrix $A$ (as one realization of $P$) has the 
resulting pairs plot:

```{r a_ase, fig.cap = 'Pair plots of the adjacency spectral embedding of $A$ Since $K=2$ is fixed, we use the signature $(3, 1)$ in this embedding.'}
pairs(Zhat, col = z, asp = 1)
```

Minimizing the objective function in (1) results in the following embedding:

```{r}
XU.hat <- transform.embedding(Zhat, z)
pairs(XU.hat, col = z, asp = 1)
```
